---
title: "Persons in Crisis Calls and Police Apprehension"
subtitle: "Exploring the Relationship Between Volume of Calls and Apprehensions Made"
author: 
  - Rosemarie Topp
thanks: "Code and data are available at: [https://github.com/rosiet1039/Persons-in-Crisis-Calls-and-Police-Apprehension](https://github.com/rosiet1039/Persons-in-Crisis-Calls-and-Police-Apprehension)."
date: today
date-format: long
abstract: "The Toronto Police Service is responsible for attending to high volume of Person in Crisis calls. There is much debate over whether or not they are the appropriate group to be dealing with situations involving mental illness and substance use. This paper provides evidence of a link between high volumes of calls and low proportions of apprehensions made across police divisions, which could indicate that over tasking police officers may lead to a mishandling of mental health crises."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(knitr)
library(readr)
library(dplyr)
library(ggplot2)
pic <- read.csv("C:/Users/owner/starter_folder/data/02-analysis_data/persons_in_crisis_data.csv")
```


# Introduction

A Person in Crisis (PIC) is defined by the Toronto Police Services (TPS) as "a member of the public whose behavior brings them into contact with emergency services, either because of an apparent need for urgent care within the mental health system, or because they are otherwise experiencing a mental, emotional or substance use crisis involving behavior that is sufficiently erratic, threatening or dangerous that emergency services are called in order to protect the person or those around them" [@citeTPS]. A subset of these individuals may be apprehended if an officer has reasonable grounds to believe they pose a danger to themself or others. Ontario's Mental Health Act (MHA) describes the criteria the PIC must fit in order to be brought into custody [@citeMHA]. Many of these criteria require the officer to assess the body language and behavioural cues of the PIC, which can be complicated when dealing with those suffering from a mental health crisis. The use of officers in dealing with mental health crises has been widely criticized, with many mental health professionals stating that the police do not have sufficient training to properly deescalate these conflicts [@citeNasser]. There are also complaints from police, who have stated that dealing with persons with mental illness overwhelms their working hours [@citeGriffiths].

TPS uploads data on PIC calls for service that have been attended, including information such as the time and date of the event, whether or not the PIC was apprehended, and the neighbourhood of the event. Data on these calls ranging from 1 January 2014 to 12 June 2015 can be found on Open Data Toronto [@citeOpenData], and an interactive dashboard of summary statistics is on the TPS Public Safety Data Portal [@citeDashboard].

The city of Toronto is currently divided into 158 different neighbourhoods, however this paper will be looking at the 140 neighbourhood division which was in place when the data was collected [@citeNeighbourhoodProfiles]. The TPS is divided into 17 different divisions which each cover some subset of these neighbourhoods [@citeDivisions].

This paper will go through the Persons in Crisis Calls for Service Attended data set available on Open Data Toronto. @sec-data will describe the variables of interest, looking at the differences in call volume and apprehension rates across police divisions, as well as across Toronto's neighbourhoods. @sec-discussion will discuss findings in further detail, and @sec-limitations will briefly point out some limitations and ideas for further research.




# Data {#sec-data}

## Overview

The data analyzed in this paper comes from Open Data Toronto, and was accessed through the `opendatatoronto` package [@citeOpenData] [@citeOpenDataR]. All handling of data was done using the statistical programming language R [@citeR]. Packages `tidyverse` [@citeTidyverse], `readr` [@citeReadr], and `dplyr` [@citeDplyr] were used to clean, analyze, and simulate data. All plots were created using `ggplot2` [@citeGgplot2]. `knitr` was used to help format this document [@citeKnitr].

## Measurement

The Persons in Crisis Calls for Service Attended dataset contains information about every PIC call attended by an officer of the TPS from 1 January 2014 to 12 June 2015. When a call is attended, the date, day of the week, hour, and neighbourhood (in both the 140 and 158 neighbourhood systems [@citeNeighbourhoodProfiles]) of the event is recorded, as well as the TPS division that was in attendance. These events are are aggregated into three "types": Person in Crisis, Suicide-related, and Overdose, and this is also recorded. Additionally, whether or not the PIC was apprehended and whether or not an occurrence was made is noted. This paper will be focusing on three variables in particular: the TPS division, the neighbourhood, and whether or not the event resulted in an apprehension.

## Variables of Interest

### Division

The TPS is divided into 17 divisions, each of which cover some region of Toronto. While most divisions attended around 1,250 to 2,000 calls over the recorded time period, divisions 14 and 51 were particularly notable, attending 3,425 and 3,847 events respectively (@fig-divisions). These divisions are two of the smallest area-wise, but both cover prominent downtown neighbourhoods like Moss Park, which has a reputation for being one of the roughest areas in the city [@citeDivisions] [@citeTL]. 

```{r}
#| label: fig-divisions
#| fig-cap: PIC Calls for Service Attended by TPS Divisions
#| echo: false

ggplot(pic, aes(x = division)) +
  geom_bar(fill = "#393e38") +
  theme_minimal() +
  theme(
        panel.grid.major.x = element_blank(),
        plot.background = element_rect(fill = "#f8f6f5"),
        legend.position = "none",
        plot.title = element_text(colour = "#393e38", family = "serif", vjust = 1),
        axis.title = element_text(colour = "#393e38", family = "serif"),
        legend.title = element_text(colour = "#393e38", family = "serif"),
        axis.text.x = element_text(colour = "#989394", vjust = 5),
        axis.text.y = element_text(colour = "#989394", hjust = 1)
        ) +
  labs(title = "Number of Calls Attended by Division",
       x = "Division", y = "Number of Calls Attended") +
  annotate("text", x = 4, y = 3550, label = nrow(filter(pic, division=="D14")), colour = "#393e38", size = 3) +
  annotate("text", x = 13, y = 4000, label = nrow(filter(pic, division=="D51")), colour = "#393e38", size = 3)
```

### Neighbourhoods

From the late 1990s to very recently in 2022, Toronto was divided into 140 neighbourhoods. Due to an increase in the population density in certain regions, the city made the decision to split some neighbourhoods creating a total of 158 [@citeNeighbourhoodProfiles]. Since the data in question was collected between 2014 and 2015, this paper will use the 140 neighbourhood system. @fig-neighbourhoods shows the top three neighbourhoods with the highest number of PIC calls.

```{r}
#| label: fig-neighbourhoods
#| fig-cap: PIC Calls for Service Attended by Neighbourhoods
#| fig-subcap: ¹73 - Moss Park (Division 51), 75 - Church-Yonge Corridor (Division 51), 76 - Bay Street Corridor (Division 52).
#| echo: false

pic |>
  group_by(hood_140) |>
  summarize(n_calls = n()) |>
  top_n(3, n_calls) |>
  kable(
    "latex",
    booktabs = TRUE,
    col.names = c("¹Neighbourhood", "Number of PIC Calls"))
```


*Neighbourhood names: 73 - Moss Park (Division 51), 75 - Church-Yonge Corridor (Division 51), 76 - Bay Street Corridor (Division 52).

### Apprehension Made

A subset of the PIC calls result in the person in question being apprehended, usually to be admitted to a psychiatric facility [@citeTPS]. This occurs when the officer believes the person in crisis poses a danger to themself or others. Overall, `r round(100 * mean(pic$apprehension_made == "Yes"), 1)`% of the calls resulted in an apprehension.

There appears to be a negative relationship between the number of calls overall and the proportion of calls resulting in an apprehension, both when grouping by neighbourhood @-neighbourhood_apprehensions and by division @fig-divisions_apprehensions. We can see that division 51, which attended the highest overall number of calls, has the lowest proportion of apprehensions made. Conversely, division 42 has the lowest calls attended and the highest proportion of apprehensions.

There could be confounding variables involved in this relationship, and more research should be done on this front.

```{r}
#| label: fig-divisions_apprehensions
#| fig-cap: Number of Calls v Proportion of Calls with Apprehension Made by Division
#| echo: false
#| message: false

pic |>
  group_by(division) |>
  summarize(prop_yes = mean(apprehension_made == "Yes"), n_calls = n()) |>
  ggplot(aes(x = n_calls, y = prop_yes)) +
  geom_point(colour = "#393e38") +
  geom_smooth(method = "lm", se = FALSE, colour = "#989394") +
  theme_minimal() +
  theme(panel.grid.minor.x = element_blank(),
        plot.background = element_rect(fill = "#f8f6f5"),
        legend.position = "none",
        plot.title = element_text(colour = "#393e38", family = "serif", vjust = 1),
        axis.title = element_text(colour = "#393e38", family = "serif"),
        legend.title = element_text(colour = "#393e38", family = "serif"),
        axis.text.x = element_text(colour = "#989394", vjust = 10),
        axis.text.y = element_text(colour = "#989394", hjust = 1)) +
  labs(title = "Number of Calls v Proportion of Calls with Apprehension Made by Division",
       x = "Number of Calls", y = "Proportion with Apprehension Made") +
  annotate("text", x = 1250, y = 0.3005, label = "D42", colour = "#393e38", size = 3) +
  annotate("text", x = 3425, y = 0.195, label = "D14", colour = "#393e38", size = 3) +
  annotate("text", x = 3847, y = 0.155, label = "D51", colour = "#393e38", size = 3)
```

```{r}
#| label: fig-neighbourhood_apprehensions
#| fig-cap: Number of Calls v Proportion of Calls with Apprehension Made by Neighbourhood
#| echo: false
#| message: false

pic |>
  group_by(hood_140) |>
  summarize(prop_yes = mean(apprehension_made == "Yes"), n_calls = n()) |>
  ggplot(aes(x = n_calls, y = prop_yes)) +
  geom_point(fill = "#393e38") +
  geom_smooth(method = "lm", se = FALSE, colour = "#989394") +
  theme_minimal() +
  theme(panel.grid.minor.x = element_blank(),
        plot.background = element_rect(fill = "#f8f6f5"),
        legend.position = "none",
        plot.title = element_text(colour = "#393e38", family = "serif", vjust = 1),
        axis.title = element_text(colour = "#393e38", family = "serif"),
        legend.title = element_text(colour = "#393e38", family = "serif"),
        axis.text.x = element_text(colour = "#989394", vjust = 10),
        axis.text.y = element_text(colour = "#989394", hjust = 1)) +
  labs(title = "Number of Calls v Proportion of Calls with Apprehension Made by Neighbourhood",
       x = "Number of Calls", y = "Proportion with Apprehension Made") +
    annotate("text", x = 1400, y = 0.16, label = "Church-Yonge Corridor (75)", colour = "#393e38", size = 3) +
    annotate("text", x = 1350, y = 0.125, label = "Moss Park (73)", colour = "#393e38", size = 3)
```

\newpage

## Discussion {#sec-discussion}

There is much debate over whether or not police should be responsible for apprehending those experiencing a mental health crisis. Many have stated that these apprehensions are deeply traumatic and further escalate a situation in which the person in crisis has no real intent to cause harm [@citeRandall]. Moreover, police involvement in a crisis situation can even result in unnecessary death [@citeAghasi]. That said, many feel the lives of unarmed mental health professionals would be at risk were they to enter a situation with a violent individual in crisis [@citeNasser].

Whether or not the TPS should be handling these crises, the proportion of apprehension should ideally be approximately the same across neighbourhoods and divisions. The negative relationship between this proportion and the total volume of calls could indicate an overwhelmed and over tasked police force that simply does not have the time to deal with mental health crises in the appropriate manner. According to the Mental Health Act, officers must stay with an apprehended individual until they are admitted into a psychiatric facility [@citeMHA]. This takes up a significant proportion of their time; a study on the Edmonton police force found that over a three month period, officers spent around 1,500 hours in hospitals, approximately 54% of which were spent dealing with persons with mental illness [@citeGriffiths]. Across Canada, about one in five encounters with police involve persons struggling with mental illness or substance use [@citeGoC].

This data provides further evidence that the Toronto Police Services may not be the appropriate group to handle Persons in Crisis. Not only has this delegation been shown to further traumatize those involved, often escalating situations to dangerous levels, but it also takes up police hours that could be spent on situations better suited for armed officers. It is particularly concerning if this over tasking is causing the police to mishandle mental health crises.

## Limitations and Further Research {#sec-limitations}

There could be confounding variables involved in the relationships analyzed in this paper. In particular, this analysis did not look at Neighbourhood Profile data, which can be found on Open Data Toronto. Variables such as neighbourhood income, non-english languages spoken, and race could play a significant role. Further research should be done in order to better understand the causes of police apprehension in PIC situations.


\newpage


# References


